---
title: "Time Series Analysis of *Aedes albopictus* Eggs"
author: "Daniele Da Re, Andrea Radici"
date: "2023-07-14"
output: 
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    theme: readable
---

```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
opts_knit$set(root.dir = "/home/dared/GitHub/albopictus_lagged_analysis/")
```

## 0. Introduction

This report presents a time series analysis of *Aedes albopictus* egg abundance using the **VectAbundance dataset** ([Da Re et al., 2024](https://doi.org/10.1038/s41597-024-03482-y)). The study focuses on understanding the influence of climatic variables on egg dynamics, model validation, and climate change projections.

The specific objectives of this study are:


* **Climatic Drivers**: Investigate the effect of lagged temperature and precipitation—individually and in combination—on egg abundance dynamics.

* **Model Validation and Prediction**: Apply a Distributed Lag Non-Linear Model (DLNM), calibrated on the complete VectAbundance dataset across all locations, to:

* **Predict egg abundance during missing weeks**, allowing estimation of the start and end of the oviposition season.

* **Validate predictions on independent locations**: Montpellier (France) dataset (still to do).

* **Climate Change Simulation (optional)**: Simulate future oviposition dynamics under climate change projections.

## 1. Preprocessing

Load packages and observations. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(Epi) 
library(dlnm) 
library(tsModel)
library(splines) 
library(mgcv)
library(gnm) 
library(pbs) 
library(terra)
library(sf)
library(mapview)

bio.matrix_sel <- readRDS("VectAbundance015_ER5.RDS") %>% 
  select(ID, Country, Region, year, latitude, longitude, week, date, value, 
         tas, tp) %>% 
  rename(eggs=value, medianTweek= tas, cumPrecweek=tp, long= longitude, lat = latitude) %>% 
  mutate(month=month(date), 
         eggs=floor(eggs)) %>% 
  as_tibble()
bio.matrix_sel
```

A map was created to illustrate the geographical distribution of the sampling locations.

```{r}
bio.matrix_sel %>% 
  select(long, lat) %>% 
  distinct() %>% 
  st_as_sf(coords=c("long", "lat"), crs=4326) %>% 
  mapview()
```
One location ID was excluded from the initial analysis to be used later for model validation.
```{r}
ID_val = "6776"

mydf <- bio.matrix_sel %>% 
   filter(ID != ID_val) 

# plot validation plot location
bio.matrix_sel %>% 
  filter(ID == ID_val) %>%  
  select(long, lat) %>% 
  distinct() %>% 
  st_as_sf(coords=c("long", "lat"), crs=4326) %>% 
  mapview()
```

Exploratory plots were generated to visualize the response variable (egg abundance) across the different regions.

```{r}
plot(eggs~date,  data=mydf, cex=0.05 ,  xlab="Date", ylab="Number of eggs", bty="l")
summary(mydf$eggs)
par(mfrow=c(1,2))
hist(mydf$eggs, main = "Histogram of Number of eggs")
hist(log10(mydf$eggs + 1 ), main = "Histogram of log10(Number of eggs +1)")
par(mfrow=c(1,1))
```

Inspect an individual time series ID_subset

```{r}
ID_subset = "9347"

dsub <- subset(mydf, ID==ID_subset) 

par(mar=c(2,3,1,1))
layout(matrix(1:3, nrow=3), heights=c(2,2,2))
plot(eggs~date, data=dsub, type="s", lty=2, bty="l", xlab="",
     ylab="Number of eggs", mgp=c(2.2,0.7,0), lab=c(5,3,7))
plot(medianTweek~date, data=dsub, type="h", lty=1, bty="l", col=2, xlab="",
     ylab="Temperature  (°C)")
plot(cumPrecweek~date, data=dsub, type="h", lty=1, bty="l", col=4, xlab="Date",
     ylab="Cumulative precipitation (mm)")
layout(1)
par(mar=c(5,4,4,1)+0.1)
```

We apply log10 transformation to **brutally** account for the negative binomial distribution
```{r}
mydf$eggsL10<- log10(mydf$eggs+1)
```

Investigate the latitudinal variation
```{r}
plot(unique(mydf$lat), ylim=c(40,50), xaxt='n', xlab="", ylab="Latitude",  lty=1, bty="n")
```

## 2. Trend and Seasonality

Define spline transformation to capture the seasonal trends.
```{r}
# long interannual temporal trend
dftrend <- round(as.numeric(diff(range(mydf$date))/365.25 * length(unique(mydf$year)))) 
btrend <- ns(mydf$date, knots=equalknots(mydf$date, dftrend-1))
# seasonal intraannual trend
mydf$week <- as.numeric(mydf$week)
mydf$year <- as.factor(mydf$year)
bseas <- ns(mydf$week, df=4) 
```

Define the stratum as a grouping factor (so far, we're not using it)
```{r}
mydf$stratum <- with(mydf, year)
```

## 3. Univariate models

### 3.1 Temperature

To model the delayed and potentially non-linear effect of temperature on egg abundance, we define a crossbasis using a Distributed Lag Non-Linear Model (DLNM) framework.

A B-spline is applied to the temperature dimension, with knots positioned at the lower and upper quartiles of the temperature distribution. This allows flexible modeling of the non-linear relationship between temperature and the response.

For the lag dimension, we consider effects up to 8 weeks. A spline is also used here, with knots placed at the 1st and 4th week, capturing both short- and medium-term lagged associations.

```{r}
tknots<-unname(quantile(mydf$medianTweek, probs = c(0.25, 0.75), na.rm = TRUE))
#original

cbTemp <- crossbasis(mydf$medianTweek, lag=8,
                     argvar=list(fun="bs", degree=2, knots=tknots),
                     arglag=list(knots=c(1,4)),
                     group=mydf$Region)
```
   
We now run the model suite, consisting of the following univariate configurations:

* **mod_T**: A univariate model including only temperature as the predictor, used to assess the direct effect of temperature on the response variable.

* **mod_T.seas**: A univariate model including both temperature and a seasonal component, to evaluate the temperature effect while accounting for underlying seasonal trends.

* **mod_T_pred**: A predictive univariate model including temperature, in which the crossbasis is pre-computed and passed as a separate data frame. This facilitates predictions on new data or for missing time points.

```{r}
# Fit mod_Tel to analyse coefficients
mod_T <- gnm(eggsL10 ~ cbTemp,
           eliminate=stratum, #grouping factor
           data=mydf,
           family=gaussian(),
           na.action= "na.exclude")

mod_T.seas <- gnm(eggsL10 ~ cbTemp+ bseas*year, # + btrend,
           eliminate=stratum, #grouping factor
           data=mydf,
           family=gaussian(),
           na.action= "na.exclude")

# define predictive model
cbTemp_df <- as.data.frame(cbTemp)
colnames(cbTemp_df) <- paste0("cb", seq_len(ncol(cbTemp_df)))

mydf_ext <- cbind(mydf, cbTemp_df)

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames(cbTemp_df), collapse = " + ")))

# Fit mod_Tel to predict on new data
mod_T_pred <- gnm(form,
                data = mydf_ext,
                family = gaussian(),
                na.action = "na.exclude")
bseas_df <- as.data.frame(bseas)
colnames(bseas_df) <- paste0("bt", seq_len(ncol(bseas_df)))
mydf_ext.seas <- cbind(mydf, cbTemp_df, bseas_df) #btrend

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames( cbind(cbTemp_df,  bseas_df)), collapse = " + "))) # btrend_df,

mod_T.seas_pred <- gnm(form,
                data = mydf_ext.seas,
                family = gaussian(),
                na.action = "na.exclude")
```
Inspect the AIC for the univariate model with temperature and the model with temperature and seasonality

```{r}
AIC(mod_T, mod_T.seas)
```

#### 3.1.1 Response curves

Investigate the response curves on temperature.

```{r}
#centering point at 10°C
myCen <- 10
cptmean <- crosspred(cbTemp, mod_T, cen=myCen, by=1, at=0:40)
cptmean.seas <- crosspred(cbTemp, mod_T.seas, cen=myCen, by=1, at=0:40)

# Overall cumulative
plot(cptmean, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature + Seasonality: overall cumulative exposure-response")
```

Temperature and lags together

```{r}
# Contour map 2D
plot(cptmean, "contour", xlab="Temperature (°C)", ylab="Lag", 
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, "contour", xlab="Temperature (°C)", ylab="Lag", 
     main="Temperature + seasonality: overall cumulative exposure-response")
```

```{r}
# 3D surface
plot(cptmean, xlab="Temperature (°C)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Temperature: exposure-lag-response", col=2)
plot(cptmean.seas, xlab="Temperature (°C)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Temperature + seas : exposure-lag-response", col=2)
```

Plot the effect at different lags (univariate model only)

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, var=5, xlab="Lag", ylab="Increment (Log10)", col=2, 
     main="Temperature at 5 °C")

plot(cptmean, var=15, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 15 °C")

plot(cptmean, var=25, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 25 °C")

plot(cptmean, var=40, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 40 °C")
layout(1)
```

Plot the effect at different Temperatures (univariate model only)

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, lag=0, xlab="T", ylab="Increment (Log10)", col=2, 
     main="lag=0")

plot(cptmean, lag=2, xlab="T", ylab="Increment (Log10)", col=2, 
     main="lag=2")

plot(cptmean, lag=4, xlab="T", ylab="Increment (Log10)", col=2,
     main="lag=4")

plot(cptmean, lag=8, xlab="T", ylab="Increment (Log10)", col=2,
     main="lag=8")
layout(1)
```
#### 3.1.2 Plot fitted values

Plot values predicted on one random location selected from the calibration dataset.

```{r}
mydf$fitted <- 10^(fitted(mod_T_pred))-1 
mydf$fitted.seas <- 10^(fitted(mod_T.seas_pred))-1 

mydf_temp <- mydf %>%
  filter(ID == ID_subset)

plot(mydf_temp$date, mydf_temp$eggs, type="p" ,col="blue", ylab="Y", main= "Temperature only model")
lines(mydf_temp$date, mydf_temp$fitted, col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_temp$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_temp$eggs, mydf_temp$fitted)$estimate,3)))


plot(mydf_temp$date, mydf_temp$eggs, type="p" ,col="blue", ylab="Y", main= "Temperature + seas model")
lines(mydf_temp$date, mydf_temp$fitted.seas, col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T + Seas"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_temp$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_temp$eggs, mydf_temp$fitted.seas)$estimate,3)))

```

#### 3.1.3 Validation plot

Plot values predicted on the location that was previously excluded from the model calibration.

```{r}
mydf_val <- bio.matrix_sel %>% 
  filter(ID == ID_val) %>% 
  mutate(week=as.integer(week), 
         year=as.factor(year))

regions <- unique(mydf$Region)

# Build newdf with same shape as original (if needed)
newdf <- data.frame(date = mydf_val$date,
                    medianTweek = mydf_val$medianTweek)

# Build new crossbasis
cbTemp_val <- crossbasis(newdf$medianTweek, lag=8,
                     argvar=list(fun="bs", degree=2, knots=tknots),
                     arglag=list(knots=c(1,4)),
                     group=newdf$Region)

cbTemp_val_df <- as.data.frame(cbTemp_val)
colnames(cbTemp_val_df) <- paste0("cb", seq_len(ncol(cbTemp_val_df)))
newdf_ext <- cbind(newdf, cbTemp_val_df)

# Predict
log10eggsPred <- predict(mod_T_pred, newdata = newdf_ext, type = "response")
mydf_val$eggsPred  = 10^(log10eggsPred)-1

#plot
plot(mydf_val$date, mydf_val$eggs, col = "blue")
lines(newdf$date, mydf_val$eggsPred , col = "darkgreen", lty=2)
# plot(newdf$date, predY, col = "darkgreen")
legend("topright", legend=c("observed eggs", "predicted eggs"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_val$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_val$eggs, mydf_val$eggsPred)$estimate,3)))
```

Predict also using the model with both Temperature and seasonality.

```{r}
# long interannual temporal trend
btrend <- ns(mydf_val$date, knots=equalknots(mydf_val$date, dftrend-1))
# seasonal intraannual trend
bseas_pred <- ns(mydf_val$week, df=4) #week

bseas_pred_df <- as.data.frame(bseas_pred)
colnames(bseas_pred_df) <- paste0("bt", seq_len(ncol(bseas_pred_df)))

mydf_ext.seas <- cbind(newdf, cbTemp_val_df, bseas_pred_df)

# Predict
log10eggsPred <- predict(mod_T.seas_pred, newdata = mydf_ext.seas, type = "response")
mydf_val$eggsPred_seas  = 10^(log10eggsPred)-1

# plot
plot(mydf_val$date, mydf_val$eggs, type="p" ,col="blue", ylab="Y", main= "Temperature + seas model")
lines(mydf_val$date, mydf_val$eggsPred_seas , col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T + Seas"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_val$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_val$eggs, mydf_val$eggsPred_seas )$estimate,3)))
```

#### 3.1.4 Risk factors

```{r}
#Estimation at different T and lags
t.risk.v <- as.character(5*(0:(40/5)))

t.risk.df = data.frame(temp = t.risk.v,
                       allfit = NA,
                       alllow = NA,
                       allhigh = NA)

# can do it both with cptmean and cptmean.seas
for (t.risk in t.risk.v){
  t.risk.df[which(t.risk== t.risk.v),2:4] = 
    round(with(cptmean.seas, cbind(allfit, alllow, allhigh)[t.risk,]), 2)
}

t.risk.df 

# lag.risk.v <- as.character(0:8)
# 
# lag.risk.df = data.frame(lag = lag.risk.v,
#                        allfit = NA,
#                        alllow = NA,
#                        allhigh = NA)
# 
# for (lag.risk in lag.risk.v){
#   lag.risk.df[which(lag.risk== lag.risk.v),2:4] = 
#     round(with(cptmean, cbind(allfit, alllow, allhigh)[,lag.risk]), 2)
# }

```

### 3.2. Precipitation

To assess the delayed and non-linear effects of weekly cumulative precipitation on egg abundance, we define a crossbasis function using the Distributed Lag Non-Linear Model (DLNM) approach.

* For the exposure dimension (cumPrecweek), we apply a B-spline of degree 2 with knots placed at the 50th (median) and 90th percentiles of the precipitation distribution. This allows the model to flexibly capture non-linear relationships between precipitation and the response variable.

* For the lag dimension, we consider a maximum lag of 8 weeks, with internal knots at weeks 1 and 4. This structure captures both short-term and delayed effects. 

```{r}
pknots<-unname(quantile(mydf$cumPrecweek, probs = c(0.5, 0.90), na.rm = TRUE))
#original
cbPrec <- crossbasis(mydf$cumPrecweek, lag=8,
                     argvar=list(fun="bs", degree=2, knots=pknots),
                     arglag=list(knots=c(1,4)),
                     group=mydf$Region)
```

We now run a series of univariate models to investigate the effect of precipitation on egg abundance. The model suite includes the following configurations:

* **mod_P**: A univariate model including only precipitation as the predictor, used to evaluate the direct effect of precipitation on the response variable.

* **mod_P.seas**: A univariate model including both precipitation and a seasonal component, allowing us to assess the precipitation effect while adjusting for seasonal patterns.

* **mod_P_pred**: A predictive univariate model including precipitation, where the crossbasis is pre-computed and passed as a separate data frame. This setup supports forecasting or imputing values in weeks with missing data.

```{r}
# Fit mod_Pel to analyse coefficients
mod_P <- gnm(eggsL10 ~ cbPrec,
             #eliminate=stratum, #grouping factor
             data=mydf,
             family=gaussian(),
             na.action= "na.exclude")

mod_P.seas <- gnm(eggsL10 ~ cbPrec+ bseas, # + btrend,
                  #eliminate=stratum, #grouping factor
                  data=mydf,
                  family=gaussian(),
                  na.action= "na.exclude")

# define predictive model
cbPrec_df <- as.data.frame(cbPrec)
colnames(cbPrec_df) <- paste0("cb", seq_len(ncol(cbPrec_df)))

mydf_ext <- cbind(mydf, cbPrec_df)

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames(cbPrec_df), collapse = " + ")))

# Fit mod_Pel to predict on new data
mod_P_pred <- gnm(form,
                  data = mydf_ext,
                  family = gaussian(),
                  na.action = "na.exclude")

bseas_df <- as.data.frame(bseas)
colnames(bseas_df) <- paste0("bt", seq_len(ncol(bseas_df)))
mydf_ext.seas <- cbind(mydf, cbPrec_df, bseas_df) #btrend

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames( cbind(cbPrec_df, bseas_df)), collapse = " + "))) # btrend_df,

mod_P.seas_pred <- gnm(form,
                       data = mydf_ext.seas,
                       family = gaussian(),
                       na.action = "na.exclude")
```

Inspect the AIC for the univariate model with precipitation and the model with precipitation and seasonality

```{r}
AIC(mod_P, mod_P.seas)
```

#### 3.2.1 Response curves

```{r}
#centering point at 25 mm
myCen <- 25
cppcum <- crosspred(cbPrec, mod_P, cen=myCen, by=1, at=0:150)
cppcum.seas <- crosspred(cbPrec, mod_P.seas, cen=myCen, by=1, at=0:150)

# Overall cumulative
plot(cppcum, ptype="overall", xlab="Precipitation (mm)", ylab="Increment (Log10)", col=2,
     main="Precipitation: overall cumulative exposure-response")
plot(cppcum.seas, ptype="overall", xlab="Precipitation (mm)", ylab="Increment (Log10)", col=2,
     main="Precipitation + Seasonality: overall cumulative exposure-response")
```

```{r}
# Contour map 2D
plot(cppcum, "contour", xlab="Precipitation (mm)", ylab="Lag", 
     main="Precipitation: overall cumulative exposure-response")
plot(cppcum.seas, "contour", xlab="Precipitation (mm)", ylab="Lag", 
     main="Precipitation + seas: overall cumulative exposure-response")
```

```{r}
# 3D surface
plot(cppcum, xlab="Precipitation (mm)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Precipitation: exposure-lag-response", col=2)
plot(cppcum.seas, xlab="Precipitation (mm)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Precipitation + seas : exposure-lag-response", col=2)
```

Plot the effect at different lags

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cppcum, var=0, xlab="Lag", ylab="Increment (Log10)", col=2, 
     main="Precipitation at 0 mm")

plot(cppcum, var=10, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 10 mm")

plot(cppcum, var=50, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 50 mm")

plot(cppcum, var=100, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 100 mm")
layout(1)
```

Plot the effect at different precipitation values

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cppcum, lag=0, xlab="P", ylab="Increment (Log10)", col=2, 
     main="lag=0")

plot(cppcum, lag=2, xlab="P", ylab="Increment (Log10)", col=2, 
     main="lag=2")

plot(cppcum, lag=4, xlab="P", ylab="Increment (Log10)", col=2,
     main="lag=4")

plot(cppcum, lag=8, xlab="P", ylab="Increment (Log10)", col=2,
     main="lag=8")
layout(1)
```

#### 3.2.2 Plot fitted values

Plot values predicted on one random location selected from the calibration dataset.

```{r}
mydf$fitted <- 10^(fitted(mod_P_pred))-1 
mydf$fitted.seas <- 10^(fitted(mod_P.seas_pred))-1 

mydf_temp <- mydf %>%
  filter(ID == ID_subset)

plot(mydf_temp$date, mydf_temp$eggs, type="p" ,col="blue", ylab="Y", main= "Precipitation only model")
lines(mydf_temp$date, mydf_temp$fitted, col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_temp$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_temp$eggs, mydf_temp$fitted)$estimate,3)))


plot(mydf_temp$date, mydf_temp$eggs, type="p" ,col="blue", ylab="Y", main= "Precipitation + seas model")
lines(mydf_temp$date, mydf_temp$fitted.seas, col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T + Seas"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_temp$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_temp$eggs, mydf_temp$fitted.seas)$estimate,3)))

```

#### 3.2.3 Validation plot

Plot values predicted on the location that was previously excluded from the model calibration.

```{r}
mydf_val <- bio.matrix_sel %>% 
  filter(ID == ID_val)

regions <- unique(mydf$Region)

# Build newdf with same shape as original (if needed)
newdf <- data.frame(date = mydf_val$date,
                    cumPrecweek = mydf_val$cumPrecweek)

# Build new crossbasis
cbPrec_val <- crossbasis(newdf$cumPrecweek, lag=8,
                         argvar=list(fun="bs", degree=2, knots=tknots),
                         arglag=list(knots=c(1,4)),
                         group=newdf$Region)

cbPrec_val_df <- as.data.frame(cbPrec_val)
colnames(cbPrec_val_df) <- paste0("cb", seq_len(ncol(cbPrec_val_df)))
newdf_ext <- cbind(newdf, cbPrec_val_df)

# Predict
log10eggsPred <- predict(mod_P_pred, newdata = newdf_ext, type = "response")
mydf_val$eggsPred  = 10^(log10eggsPred)-1

plot(mydf_val$date, mydf_val$eggs, col = "blue")
lines(newdf$date, mydf_val$eggsPred , col = "darkgreen")
# plot(newdf$date, predY, col = "darkgreen")
legend("topright", legend=c("observed eggs", "predicted eggs"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_val$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_val$eggs, mydf_val$eggsPred)$estimate,3)))

```

Predict also using the model with both Temperature and seasonality.

```{r}
# long interannual temporal trend
btrend <- ns(mydf_val$date, knots=equalknots(mydf_val$date, dftrend-1))
# seasonal intraannual trend
bseas_pred <- ns(mydf_val$month, df=4) #week

bseas_pred_df <- as.data.frame(bseas_pred)
colnames(bseas_pred_df) <- paste0("bt", seq_len(ncol(bseas_pred_df)))

mydf_ext.seas <- cbind(newdf, cbPrec_val_df, bseas_pred_df)

# Predict ()
log10eggsPred <- predict(mod_P.seas_pred, newdata = mydf_ext.seas, type = "response")
mydf_val$eggsPred_seas  = 10^(log10eggsPred)-1

#plot
plot(mydf_val$date, mydf_val$eggs, type="p" ,col="blue", ylab="Y", main= "Precipitation + seas model")
lines(mydf_val$date, mydf_val$eggsPred_seas , col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T + Seas"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_val$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_val$eggs, mydf_val$eggsPred_seas )$estimate,3)))
```

#### 3.2.4 Risk factors
```{r}
#Estimation at different T and lags
p.risk.v <- as.character(c(0, 10, 20, 40, 60, 90, 120))
p.risk.df = data.frame(prec = p.risk.v,
                       allfit = NA,
                       alllow = NA,
                       allhigh = NA)

# can do it both with cppcum and cppcum.seas
for (p.risk in p.risk.v){
  p.risk.df[which(p.risk== p.risk.v),2:4] = 
    round(with(cppcum.seas, cbind(allfit, alllow, allhigh)[p.risk,]), 2)
}
p.risk.df 
```
## 4. Mutivariate model

To jointly assess the effects of temperature and precipitation on egg abundance, we implement a set of multivariate models using crossbasis functions for both predictors.

* **mod_TP**: A multivariate model including both temperature (cbTemp) and precipitation (cbPrec) as predictors. This model evaluates the combined and independent effects of the two climatic variables on the response variable (eggsL10).

* **mod_TP.seas**: An extended version of the multivariate model that also includes a seasonal component (bseas). This allows us to disentangle the climatic effects from seasonal patterns in egg abundance.

* **mod_TP_pred**: A predictive multivariate model, where the crossbasis is pre-computed and passed as a separate data frame. This setup supports forecasting or imputing values in weeks with missing data.

```{r}
# Fit mod_Tel to analyse coefficients
mod_TP <- gnm(eggsL10 ~  cbTemp + cbPrec,
           #eliminate=stratum, #grouping factor
           data=mydf,
           family=gaussian(),
           na.action= "na.exclude")

mod_TP.seas <- gnm(eggsL10 ~ cbTemp + cbPrec + bseas*year, # + btrend,
           #eliminate=stratum, #grouping factor
           data=mydf,
           family=gaussian(),
           na.action= "na.exclude")

# define predictive model
cbTemp_df <- as.data.frame(cbTemp)
cbPrec_df <- as.data.frame(cbPrec)
colnames(cbTemp_df) <- paste0("cbT", seq_len(ncol(cbTemp_df)))
colnames(cbPrec_df) <- paste0("cbP", seq_len(ncol(cbPrec_df)))

cbTempPrec_df <- cbind(cbTemp_df, cbPrec_df)
mydf_ext <- cbind(mydf, cbTempPrec_df)

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames(cbTempPrec_df), collapse = " + ")))

# Fit mod_Tel to predict on new data
mod_TP_pred <- gnm(form,
                data = mydf_ext,
                family = gaussian(),
                na.action = "na.exclude")

bseas_df <- as.data.frame(bseas)
colnames(bseas_df) <- paste0("bt", seq_len(ncol(bseas_df)))
mydf_ext.seas <- cbind(mydf, cbTempPrec_df, bseas_df) #btrend

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames( cbind(cbTempPrec_df ,  bseas_df)), collapse = " + "))) # btrend_df,

mod_TP.seas_pred <- gnm(form,
                data = mydf_ext.seas,
                family = gaussian(),
                na.action = "na.exclude")
```

Investigate the AIC on all the model produced so far: 

```{r}
data.frame( models = c("Temperature", "Temperature + seasonality", "Precipitation",  "Precipitation + seasonality", "Temperature + Precipitation", "Temperature + Precipitation + seasonality"),
            AIC =AIC(mod_T, mod_T.seas, mod_P, mod_P.seas, mod_TP, mod_TP.seas) )
```

### 4.1 response curves: temperatures

```{r}
#centering point at 10°C
myCen <- 10
cptmean <- crosspred(cbTemp, mod_TP, cen=myCen, by=1, at=0:40)
cptmean.seas <- crosspred(cbTemp, mod_TP.seas, cen=myCen, by=1, at=0:40)

# Overall cumulative
plot(cptmean, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature + Seasonality: overall cumulative exposure-response")
```

```{r}
# Contour map 2D
plot(cptmean, "contour", xlab="Temperature (°C)", ylab="Lag", 
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, "contour", xlab="Temperature (°C)", ylab="Lag", 
     main="Temperature + seasonality: overall cumulative exposure-response")
```

```{r}
# 3D surface
plot(cptmean, xlab="Temperature (°C)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Temperature: exposure-lag-response", col=2)
plot(cptmean.seas, xlab="Temperature (°C)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Temperature +seas : exposure-lag-response", col=2)
```

Plot the effect at different lags

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, var=5, xlab="Lag", ylab="Increment (Log10)", col=2, 
     main="Temperature at 5 °C")

plot(cptmean, var=15, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 15 °C")

plot(cptmean, var=25, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 25 °C")

plot(cptmean, var=40, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 40 °C")
layout(1)
```

Plot the effect at different Temperatures 

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, lag=0, xlab="T", ylab="Increment (Log10)", col=2, 
     main="lag=0")

plot(cptmean, lag=2, xlab="T", ylab="Increment (Log10)", col=2, 
     main="lag=2")

plot(cptmean, lag=4, xlab="T", ylab="Increment (Log10)", col=2,
     main="lag=4")

plot(cptmean, lag=8, xlab="T", ylab="Increment (Log10)", col=2,
     main="lag=8")
layout(1)
```
### 4.2 response curves: precipitation
```{r}
#centering point at 25 mm
myCen <- 25
cppcum <- crosspred(cbPrec, mod_TP, cen=myCen, by=1, at=0:150)
cppcum.seas <- crosspred(cbPrec, mod_TP.seas, cen=myCen, by=1, at=0:150)

# Overall cumulative
plot(cppcum, ptype="overall", xlab="Precipitation (mm)", ylab="Increment (Log10)", col=2,
     main="Precipitation: overall cumulative exposure-response")
plot(cppcum.seas, ptype="overall", xlab="Precipitation (mm)", ylab="Increment (Log10)", col=2,
     main="Precipitation + Seasonality: overall cumulative exposure-response")
```

```{r}
# Contour map 2D
plot(cppcum, "contour", xlab="Precipitation (mm)", ylab="Lag", 
     main="Precipitation: overall cumulative exposure-response")
plot(cppcum.seas, "contour", xlab="Precipitation (mm)", ylab="Lag", 
     main="Precipitation + seas: overall cumulative exposure-response")
```

```{r}
# 3D surface
plot(cppcum, xlab="Precipitation (mm)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Precipitation: exposure-lag-response", col=2)
plot(cppcum.seas, xlab="Precipitation (mm)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Precipitation + seas : exposure-lag-response", col=2)
```

Plot the effect at different lags

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cppcum, var=0, xlab="Lag", ylab="Increment (Log10)", col=2, 
     main="Precipitation at 0 mm")

plot(cppcum, var=10, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 10 mm")

plot(cppcum, var=50, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 50 mm")

plot(cppcum, var=100, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 100 mm")
layout(1)
```

Plot the effect at different precipitation values

```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cppcum, lag=0, xlab="P", ylab="Increment (Log10)", col=2, 
     main="lag=0")

plot(cppcum, lag=2, xlab="P", ylab="Increment (Log10)", col=2, 
     main="lag=2")

plot(cppcum, lag=4, xlab="P", ylab="Increment (Log10)", col=2,
     main="lag=4")

plot(cppcum, lag=8, xlab="P", ylab="Increment (Log10)", col=2,
     main="lag=8")
layout(1)
```

### 4.3 Plot fitted values

Plot values predicted on one random location selected from the calibration dataset.

```{r}
mydf$fitted <- 10^(fitted(mod_TP_pred))-1 
mydf$fitted.seas <- 10^(fitted(mod_TP.seas_pred))-1 

mydf_temp <- mydf %>%
  filter(ID == ID_subset)

plot(mydf_temp$date, mydf_temp$eggs, type="p" ,col="blue", ylab="Y", main= "T + P only model")
lines(mydf_temp$date, mydf_temp$fitted, col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T + P"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_temp$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_temp$eggs, mydf_temp$fitted)$estimate,3)))

plot(mydf_temp$date, mydf_temp$eggs, type="p" ,col="blue", ylab="Y", main= "T + P + seas model")
lines(mydf_temp$date, mydf_temp$fitted.seas, col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted T + P + Seas"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_temp$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_temp$eggs, mydf_temp$fitted.seas)$estimate,3)))

```

### 4.4 Validation plot 

Plot values predicted on the location that was previously excluded from the model calibration.

```{r}
mydf_val <- bio.matrix_sel %>% 
  filter(ID == ID_val)

regions <- unique(mydf$Region)

# Build newdf with same shape as original (if needed)
newdf <- data.frame(date = mydf_val$date,
                    cumPrecweek = mydf_val$cumPrecweek,
                    medianTweek = mydf_val$medianTweek)

# Build new crossbasis (T)
cbTemp_val <- crossbasis(newdf$medianTweek, lag=8,
                     argvar=list(fun="bs", degree=2, knots=tknots),
                     arglag=list(knots=c(1,4)),
                     group=newdf$Region)

# Build new crossbasis (P)
cbPrec_val <- crossbasis(newdf$cumPrecweek, lag=8,
                         argvar=list(fun="bs", degree=2, knots=tknots),
                         arglag=list(knots=c(1,4)),
                         group=newdf$Region)

cbTemp_val_df <- as.data.frame(cbTemp_val)
colnames(cbTemp_val_df) <- paste0("cbT", seq_len(ncol(cbTemp_val_df)))
cbPrec_val_df <- as.data.frame(cbPrec_val)
colnames(cbPrec_val_df) <- paste0("cbP", seq_len(ncol(cbPrec_val_df)))

cbTempPrec_val_df <- cbind(cbTemp_val_df, cbPrec_val_df)
newdf_ext <- cbind(newdf, cbTempPrec_val_df)

# Predict
log10eggsPred <- predict(mod_TP_pred, newdata = newdf_ext, type = "response")
mydf_val$eggsPred  = 10^(log10eggsPred)-1

plot(mydf_val$date, mydf_val$eggs, col = "blue", main= " T + P")
lines(newdf$date, mydf_val$eggsPred , col = "darkgreen", lty=2)
legend("topright", legend=c("observed eggs", "predicted eggs"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_val$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_val$eggs, mydf_val$eggsPred)$estimate,5)))
```

Predict also using the model with both Temperature and seasonality.

```{r}
# long interannual temporal trend
btrend <- ns(mydf_val$date, knots=equalknots(mydf_val$date, dftrend-1))
# seasonal intraannual trend
bseas_pred <- ns(mydf_val$month, df=4) #week
bseas_pred_df <- as.data.frame(bseas_pred)
colnames(bseas_pred_df) <- paste0("bt", seq_len(ncol(bseas_pred_df)))
mydf_ext.seas <- cbind(newdf, cbTempPrec_val_df, bseas_pred_df)

# Predict
log10eggsPred <- predict(mod_TP.seas_pred, newdata = mydf_ext.seas, type = "response")
mydf_val$eggsPred_seas  = 10^(log10eggsPred)-1

plot(mydf_val$date, mydf_val$eggs, type="p" ,col="blue", ylab="Y", main= " T + P + seas model")
lines(mydf_val$date, mydf_val$eggsPred_seas , col="darkgreen", lty=2)
legend("topright", legend=c("Observed", "Fitted"), col=c("blue", "darkgreen"), lty=c(1,2))
text(x=mydf_val$date[100],y=500, labels=paste("Pearson's r:", round(cor.test(mydf_val$eggs, mydf_val$eggsPred_seas )$estimate,5)))
```

### 4.5 Risk factors
```{r}
#Estimation at different T and lags
t.risk.v <- as.character(5*(0:(40/5)))

t.risk.df = data.frame(temp = t.risk.v,
                       allfit = NA,
                       alllow = NA,
                       allhigh = NA)

# can do it both with cptmean and cptmean.seas
for (t.risk in t.risk.v){
  t.risk.df[which(t.risk== t.risk.v),2:4] = 
    round(with(cptmean.seas, cbind(allfit, alllow, allhigh)[t.risk,]), 2)
}

t.risk.df 

p.risk.v <- as.character(c(0, 10, 20, 40, 60, 90, 120))

p.risk.df = data.frame(prec = p.risk.v,
                       allfit = NA,
                       alllow = NA,
                       allhigh = NA)

# can do it both with cppcum and cppcum.seas
for (p.risk in p.risk.v){
  p.risk.df[which(p.risk== p.risk.v),2:4] = 
    round(with(cppcum.seas, cbind(allfit, alllow, allhigh)[p.risk,]), 2)
}

p.risk.df 
```
## 5 Predict egg laying seasonality

In this step, we focus exclusively on the multivariate model including temperature, precipitation, and seasonality (mod_TP.seas) to estimate the seasonal pattern of egg-laying activity.

To define the start and end of the oviposition season, we adopt the following criteria based on model predictions:

* **Start of the season**: the first week in which predicted egg abundance exceeds 5 egg.

* **End of the season**: the last week in which predicted egg abundance remains above 5 egg.

This threshold-based definition allows for a consistent and interpretable estimate of the egg-laying season across different regions and years, even in the presence of missing observational data.

```{r}
pred.df <- data.frame(date=as.Date(mydf_ext.seas$date), 
                      pred = as.numeric(mydf_val$eggsPred_seas))
pred.df$year <- lubridate::year(pred.df$date)
pred.df$month <- lubridate::month(pred.df$date)
pred.df$week <- lubridate::week(pred.df$date)
pred.df$pred <- ifelse(pred.df$pred <1, 0,pred.df$pred ) 

seas <- pred.df %>% 
  group_by(year) %>% 
  filter(pred >5) %>%
  mutate(start = 1*(week==min(week)))%>%
  mutate(end = 1*(week==max(week))) %>%
  filter(start + end > 0) %>%
  ungroup()

#get mean and mode
seas %>% 
  filter(start==1) %>% 
  reframe(average.WeekStart = mean(week), 
          mode_WeekStart = names(which(table(week) == max(table(week)))))
```

### 5.1 Predict egg laying seasonality in Europe (Extrapolation)

```{r}
env.matrix <- readRDS("envMatrix_2008_2023.RDS") %>% 
  rename(medianTweek= tas, cumPrecweek=tp) %>% 
  filter(year > 2018) %>%
  as_tibble()

# Build newdf with same shape as original (if needed)
newdf <- data.frame(date = env.matrix$date,
                    cumPrecweek = env.matrix $cumPrecweek,
                    medianTweek = env.matrix $medianTweek)

# Build new crossbasis (T)
cbTemp_val <- crossbasis(newdf$medianTweek, lag=8,
                     argvar=list(fun="bs", degree=2, knots=tknots),
                     arglag=list(knots=c(1,4)),
                     group=newdf$Region)

# Build new crossbasis (P)
cbPrec_val <- crossbasis(newdf$cumPrecweek, lag=8,
                         argvar=list(fun="bs", degree=2, knots=tknots),
                         arglag=list(knots=c(1,4)),
                         group=newdf$Region)

cbTemp_val_df <- as.data.frame(cbTemp_val)
colnames(cbTemp_val_df) <- paste0("cbT", seq_len(ncol(cbTemp_val_df)))
cbPrec_val_df <- as.data.frame(cbPrec_val)
colnames(cbPrec_val_df) <- paste0("cbP", seq_len(ncol(cbPrec_val_df)))

cbTempPrec_val_df <- cbind(cbTemp_val_df, cbPrec_val_df)
newdf_ext <- cbind(newdf, cbTempPrec_val_df)

# Predict ()
log10eggsPred <- predict(mod_TP_pred, newdata = newdf_ext, type = "response")
env.matrix$eggsPred  = 10^(log10eggsPred)-1
env.matrix$eggsPred <- env.matrix$eggsPred*(env.matrix$eggsPred>0)
env.matrix$eggsPred <- round(env.matrix$eggsPred)

# beginning and end of the season defined by the 5 eggs threshold
seas.df <- env.matrix %>% 
  group_by(x, y, year) %>% 
  filter(eggsPred >5) %>%
  summarise(start = week[which(week==min(week))],
            end = week[which(week==max(week))])%>%
  ungroup()

# get onset, offset and season length
seas.av.df <- seas.df %>% 
  group_by(x, y) %>% 
  reframe(m.start = mean(start),
          m.end = mean(end),
          s.length =m.end- m.start)

slength <- seas.av.df %>% 
  select(x,y,s.length) %>% 
  terra::rast(type="xyz")

# Plot season length (months/years)
plot(slength/4, main = "Average season length (months) over 2018-2023")
```

## 6. Mutivariate GLM mod_TP

What if we avoid using the Log-transformed response variable and we go for a GLM?

```{r}

# Fit mod_Tel to analyse coefficients
mod_TP <- glm(eggs ~  cbTemp + cbPrec,
           #eliminate=stratum, #grouping factor
           data=mydf,
           family=quasipoisson,
           na.action= "na.exclude")

mod_TP.seas <- glm(eggs ~ cbTemp + cbPrec + bseas, # + btrend,
           #eliminate=stratum, #grouping factor
           data=mydf,
           family=quasipoisson,
           na.action= "na.exclude")

AIC(mod_TP, mod_TP.seas)

# define predictive model
cbTemp_df <- as.data.frame(cbTemp)
cbPrec_df <- as.data.frame(cbPrec)
colnames(cbTemp_df) <- paste0("cbT", seq_len(ncol(cbTemp_df)))
colnames(cbPrec_df) <- paste0("cbP", seq_len(ncol(cbPrec_df)))

cbTempPrec_df <- cbind(cbTemp_df, cbPrec_df)
mydf_ext <- cbind(mydf, cbTempPrec_df)

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames(cbTempPrec_df), collapse = " + ")))

# Fit mod_Tel to predict on new data
mod_TP_pred <- gnm(form,
                data = mydf_ext,
                family = gaussian(),
                na.action = "na.exclude")

# btrend_df <- as.data.frame(btrend)
# colnames(btrend_df) <- paste0("bt", seq_len(ncol(btrend_df)))
bseas_df <- as.data.frame(bseas)
colnames(bseas_df) <- paste0("bt", seq_len(ncol(bseas_df)))
mydf_ext.seas <- cbind(mydf, cbTempPrec_df, bseas_df) #btrend

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames( cbind(cbTempPrec_df ,  bseas_df)), collapse = " + "))) # btrend_df,

mod_TP.seas_pred <- gnm(form,
                data = mydf_ext.seas,
                family = gaussian(),
                na.action = "na.exclude")
```

### 6.1 response curves: temperatures

```{r}
#centering point at 10°C
myCen <- 10
cptmean <- crosspred(cbTemp, mod_TP, cen=myCen, by=1, at=0:40)
cptmean.seas <- crosspred(cbTemp, mod_TP.seas, cen=myCen, by=1, at=0:40)

# Overall cumulative
plot(cptmean, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature + Seasonality: overall cumulative exposure-response")
```
```{r}
# Contour map 2D
plot(cptmean, "contour", xlab="Temperature (°C)", ylab="Lag", 
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, "contour", xlab="Temperature (°C)", ylab="Lag", 
     main="Temperature + seasonality: overall cumulative exposure-response")
```
```{r}
# 3D surface
plot(cptmean, xlab="Temperature (°C)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Temperature: exposure-lag-response", col=2)
plot(cptmean.seas, xlab="Temperature (°C)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Temperature +seas : exposure-lag-response", col=2)
```

Plot the effect at different lags
```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, var=5, xlab="Lag", ylab="Increment (Log10)", col=2, 
     main="Temperature at 5 °C")

plot(cptmean, var=15, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 15 °C")

plot(cptmean, var=25, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 25 °C")

plot(cptmean, var=40, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Temperature at 40 °C")
layout(1)
```

Plot the effect at different T
```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, lag=0, xlab="T", ylab="Increment (Log10)", col=2, 
     main="lag=0")

plot(cptmean, lag=2, xlab="T", ylab="Increment (Log10)", col=2, 
     main="lag=2")

plot(cptmean, lag=4, xlab="T", ylab="Increment (Log10)", col=2,
     main="lag=4")

plot(cptmean, lag=8, xlab="T", ylab="Increment (Log10)", col=2,
     main="lag=8")
layout(1)
```

### 6.2 response curves: precipitation
Predict
```{r}
#centering point at 25 mm
myCen <- 25
cptmean <- crosspred(cbPrec, mod_TP, cen=myCen, by=1, at=0:150)
cptmean.seas <- crosspred(cbPrec, mod_TP.seas, cen=myCen, by=1, at=0:150)

# Overall cumulative
plot(cptmean, ptype="overall", xlab="Precipitation (mm)", ylab="Increment (Log10)", col=2,
     main="Precipitation: overall cumulative exposure-response")
plot(cptmean.seas, ptype="overall", xlab="Precipitation (mm)", ylab="Increment (Log10)", col=2,
     main="Precipitation + Seasonality: overall cumulative exposure-response")
```

```{r}
# Contour map 2D
plot(cptmean, "contour", xlab="Precipitation (mm)", ylab="Lag", 
     main="Precipitation: overall cumulative exposure-response")
plot(cptmean.seas, "contour", xlab="Precipitation (mm)", ylab="Lag", 
     main="Precipitation: overall cumulative exposure-response")
```
```{r}
# 3D surface
plot(cptmean, xlab="Precipitation (mm)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Precipitation: exposure-lag-response", col=2)
plot(cptmean.seas, xlab="Precipitation (mm)", zlab="Increment (Log10)", 
     ltheta=240, lphi=60, cex.axis=0.8, 
     main="Precipitation +seas : exposure-lag-response", col=2)
```
Plot the effect at different lags
```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, var=0, xlab="Lag", ylab="Increment (Log10)", col=2, 
     main="Precipitation at 0 mm")

plot(cptmean, var=10, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 10 mm")

plot(cptmean, var=50, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 50 mm")

plot(cptmean, var=100, xlab="Lag", ylab="Increment (Log10)", col=2,
     main="Precipitation at 100 mm")
layout(1)
```

Plot the effect at different P
```{r}
layout(matrix(1:4, ncol=2, nrow = 2))
plot(cptmean, lag=0, xlab="P", ylab="Increment (Log10)", col=2, 
     main="lag=0")

plot(cptmean, lag=2, xlab="P", ylab="Increment (Log10)", col=2, 
     main="lag=2")

plot(cptmean, lag=4, xlab="P", ylab="Increment (Log10)", col=2,
     main="lag=4")

plot(cptmean, lag=8, xlab="P", ylab="Increment (Log10)", col=2,
     main="lag=8")
layout(1)
```


## 7. Discussion points

+ **Trasformazione dei dati di abbondanza (uova)**: I dati sulle abbondanze di uova mostrano una distribuzione presumibilmente poissoniana, probabilmente zero-inflated. Abbiamo applicato una trasformazione log10(x + 1) per ridurre l’asimmetria e contenere l’influenza dei valori estremi.
 *→ Ha senso questa scelta metodologica in questo contesto? Oppure sarebbe preferibile un approccio basato direttamente su modelli zero-inflated o hurdle?*

+ **Variabilità tra stazioni e regioni**: Il dataset contiene molte stazioni distribuite su più regioni: questo da un lato arricchisce la base dati, ma introduce anche molto rumore.
*→ L’utilizzo del parametro stratum nei modelli consente di tenere conto di questa variabilità spaziale? Può essere considerato equivalente a un "grouping factor" o un effetto casuale (random effect) come nei GLMM?*

+ **Trend di lungo periodo e stagionalità**: Abbiamo incluso nel modello due spline naturali per rappresentare il trend di lungo periodo e la stagionalità.
*→ Ha senso mantenerli? La scelta di includerli dovrebbe basarsi esclusivamente sull’AIC del modello?*

+ **Valutazione dell'effetto cumulato del lag**
Appurato che non ha senso calcolare il relative risk per via della formulazione del modello (gaussiano su logaritmo), chi chiedevamo se fosse possibile valutare un effetto cumulato del lag sulla variabile di uscita, analogamente a come si fa per la temperatura
*→ Al di là dell'effettiva possibilità, ha senso?*

+ [Zero inflated models](\url{https://www.mycompiler.io/view/Bf879ujMKtH}) 

+ check interactions in dlnm in Day2_Session1_Demo_DLM.pdf 


