---
title: "Two-stages time series analysis of *Aedes albopictus* oviposition"
author: "Daniele Da Re, Andrea Radici"
date: "2025-09-03"
output: 
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    theme: readable
---


```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 

if (file.exists("C:/Users/2024ar003/Desktop/Alcuni file permanenti/Post_doc/Codice/local.R")){
opts_knit$set(root.dir = "C:/Users/2024ar003/Desktop/Alcuni file permanenti/Post_doc/Codice/04_Trento/GitShared_AlboLaggedAnalysis/")
} else {
opts_knit$set(root.dir = "/home/dared/GitHub/albopictus_lagged_analysis/")
}

```

## 0. Introduction

This report presents a time series analysis of *Aedes albopictus* egg abundance using the **VectAbundance dataset** ([Da Re et al., 2024](https://doi.org/10.1038/s41597-024-03482-y)). The study focuses on understanding the influence of climatic variables on egg dynamics, model validation, and climate change projections.

The specific objective of this study is to perform a two-stages analysis of the egg abundance temporal series to investigate the effect of lagged temperature and precipitation on egg-abundance analysis, first differentiating each regions and later pooling over the whole dataset.

## 1. Preprocessing

Load packages and observations. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(Epi) 
library(dlnm) 
library(tsModel)
library(splines) 
library(mgcv)
library(gnm) 
library(pbs) 
library(terra)
library(sf)
library(mapview)

bio.matrix_sel <- readRDS("VectAbundance015_ER5.RDS") %>% 
  select(ID, Country, Region, year, latitude, longitude, week, date, value, 
         tas, tp) %>% 
  rename(eggs=value, medianTweek= tas, cumPrecweek=tp, long= longitude, lat = latitude) %>% 
  mutate(month=month(date), 
         eggs=floor(eggs)) %>% 
  as_tibble()
```
A map was created to illustrate the geographical distribution of the sampling locations.

```{r}
bio.matrix_sel %>% 
  select(long, lat) %>% 
  distinct() %>% 
  st_as_sf(coords=c("long", "lat"), crs=4326) %>% 
  mapview()
```
One location ID was excluded from the initial analysis to be used later for model validation.
```{r}
ID_val = "6776"

mydf <- bio.matrix_sel %>% 
   filter(ID != ID_val) 

# plot validation plot location
bio.matrix_sel %>% 
  filter(ID == ID_val) %>%  
  select(long, lat) %>% 
  distinct() %>% 
  st_as_sf(coords=c("long", "lat"), crs=4326) %>% 
  mapview()
```
We apply log10 transformation to **brutally** account for the negative binomial distribution
```{r}
mydf$eggsL10<- log10(mydf$eggs+1)
```

## 2. Trend and Seasonality

Define spline transformation to capture the seasonal trends.
```{r}
# long interannual temporal trend
dftrend <- round(as.numeric(diff(range(mydf$date))/365.25 * length(unique(mydf$year)))) 
btrend <- ns(mydf$date, knots=equalknots(mydf$date, dftrend-1))
# seasonal intraannual trend
mydf$week <- as.numeric(mydf$week)
mydf$year <- as.factor(mydf$year)
bseas <- ns(mydf$week, df=4) 
```

Define the stratum as a grouping factor (so far, we're not using it)
```{r}
mydf$stratum <- with(mydf, year)
```

## 3. Bivariate model

To jointly assess the effects of temperature and precipitation on egg abundance, we implement a set of bivariate models using crossbasis functions for both predictors.

* **mod_TP**: A multivariate model including both temperature (cbTemp) and precipitation (cbPrec) as predictors. This model evaluates the combined and independent effects of the two climatic variables on the response variable (eggsL10).

* **mod_TP.seas**: An extended version of the multivariate model that also includes a seasonal component (bseas). This allows us to disentangle the climatic effects from seasonal patterns in egg abundance.

* **mod_TP_pred**: A predictive multivariate model, where the crossbasis is pre-computed and passed as a separate data frame. This setup supports forecasting or imputing values in weeks with missing data.

```{r}
# Fit mod_Tel to analyse coefficients
mod_TP <- gnm(eggsL10 ~  cbTemp + cbPrec,
           #eliminate=stratum, #grouping factor
           data=mydf,
           family=gaussian(),
           na.action= "na.exclude")

mod_TP.seas <- gnm(eggsL10 ~ cbTemp + cbPrec + bseas*year, # + btrend,
           #eliminate=stratum, #grouping factor
           data=mydf,
           family=gaussian(),
           na.action= "na.exclude")

# define predictive model
cbTemp_df <- as.data.frame(cbTemp)
cbPrec_df <- as.data.frame(cbPrec)
colnames(cbTemp_df) <- paste0("cbT", seq_len(ncol(cbTemp_df)))
colnames(cbPrec_df) <- paste0("cbP", seq_len(ncol(cbPrec_df)))

cbTempPrec_df <- cbind(cbTemp_df, cbPrec_df)
mydf_ext <- cbind(mydf, cbTempPrec_df)

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames(cbTempPrec_df), collapse = " + ")))

# Fit mod_Tel to predict on new data
mod_TP_pred <- gnm(form,
                data = mydf_ext,
                family = gaussian(),
                na.action = "na.exclude")

bseas_df <- as.data.frame(bseas)
colnames(bseas_df) <- paste0("bt", seq_len(ncol(bseas_df)))
mydf_ext.seas <- cbind(mydf, cbTempPrec_df, bseas_df) #btrend

# Build formula with those columns
form <- as.formula(paste("eggsL10 ~", paste(colnames( cbind(cbTempPrec_df ,  bseas_df)), collapse = " + "))) # btrend_df,

mod_TP.seas_pred <- gnm(form,
                data = mydf_ext.seas,
                family = gaussian(),
                na.action = "na.exclude")
```

Investigate the AIC on all the model produced so far: 

```{r}
data.frame( models = c("Temperature", "Temperature + seasonality", "Precipitation",  "Precipitation + seasonality", "Temperature + Precipitation", "Temperature + Precipitation + seasonality"),
            AIC =AIC(mod_T, mod_T.seas, mod_P, mod_P.seas, mod_TP, mod_TP.seas) )
```

### 4.1 response curves: temperatures

```{r}
#centering point at 10°C
myCen <- 10
cptmean <- crosspred(cbTemp, mod_TP, cen=myCen, by=1, at=0:40)
cptmean.seas <- crosspred(cbTemp, mod_TP.seas, cen=myCen, by=1, at=0:40)

# Overall cumulative
plot(cptmean, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature: overall cumulative exposure-response")
plot(cptmean.seas, ptype="overall", xlab="Temperature (°C)", ylab="Increment (Log10)", col=2,
     main="Temperature + Seasonality: overall cumulative exposure-response")
```