---
title: "Two-stages time series analysis of *Aedes albopictus* oviposition"
author: "Daniele Da Re, Andrea Radici"
date: "2025-09-03"
output: 
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    theme: readable
---


```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 

if (file.exists("C:/Users/2024ar003/Desktop/Alcuni file permanenti/Post_doc/Codice/local.R")){
opts_knit$set(root.dir = "C:/Users/2024ar003/Desktop/Alcuni file permanenti/Post_doc/Codice/04_Trento/GitShared_AlboLaggedAnalysis/")
} else {
opts_knit$set(root.dir = "/home/dared/GitHub/albopictus_lagged_analysis/")
}

```

## 0. Introduction

This report presents a time series analysis of *Aedes albopictus* egg abundance using the **VectAbundance dataset** ([Da Re et al., 2024](https://doi.org/10.1038/s41597-024-03482-y)). The study focuses on understanding the influence of climatic variables on egg dynamics, model validation, and climate change projections.

The specific objective of this study is to perform a two-stages analysis of the egg abundance temporal series to investigate the effect of lagged temperature and precipitation on egg-abundance analysis, first differentiating each regions and later pooling over the whole dataset.

## 1. Preprocessing

Load packages and observations. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(Epi) 
library(dlnm) 
library(tsModel)
library(splines) 
library(mgcv)
library(gnm) 
library(pbs) 
library(terra)
library(sf)
library(mapview)
library(metafor) 
library(scales)

bio.matrix_sel <- readRDS("VectAbundance015_ER5.RDS") %>% 
  select(ID, Country, Region, year, latitude, longitude, week, date, value, 
         tas, tp) %>% 
  rename(eggs=value, medianTweek= tas, cumPrecweek=tp, long= longitude, lat = latitude) %>% 
  mutate(month=month(date), 
         eggs=floor(eggs)) %>% 
  as_tibble()
```
Remove the NAs

```{r}
bio.matrix_sel <- bio.matrix_sel %>% 
  filter(!is.na(eggs))
```

A map was created to illustrate the geographical distribution of the sampling locations.

```{r}
bio.matrix_sel %>% 
  select(long, lat) %>% 
  distinct() %>% 
  st_as_sf(coords=c("long", "lat"), crs=4326) %>% 
  mapview()
```
We apply log10 transformation to **brutally** account for the negative binomial distribution
```{r}
bio.matrix_sel$eggsL10<- log10(bio.matrix_sel$eggs+1)
```

## 2. Simple case (glm)

## 2.1 First stage

Following "Day5_Session1_Demo_TwoStage.R", we organise the data as a list of dataframes for the first stage. First, we define regions, supposedly homogeneous in terms of (a) climate (b) sampling (c) mosquito pressure... And we split our df as a list of df (one for each region)

```{r}

regions <- as.character(unique(bio.matrix_sel$Region))
bio.list_sel <- lapply(regions, function(r) bio.matrix_sel[bio.matrix_sel$Region==r,])
names(bio.list_sel) <- regions

```

We will first explore a simple, gnl-based design, to obtain the relative risk as a function of temperature (but semi poisson)

```{r}

logRR <- logRRse <- vector("numeric", length(regions))
for(i in seq(bio.list_sel)) {
  sub <- bio.list_sel[[i]]
  
  nYears = length(unique(sub$year))# to regulate the df of the ns
  
  # Why temperatures in Veneto are negative?
  
  m <- glm(eggs ~  medianTweek + ns(date, 7*nYears) + (1 | ID),
    data=sub) #family=quasipoisson()
  
  logRR[i] <- coef(m)["medianTweek"]
  logRRse[i] <- sqrt(vcov(m)["medianTweek","medianTweek"])
}


```

## 2.2 Second stage
Therefore we pool estimates (with univariate random effects meta-analysis) and we plot the RR per region and divided

```{r}

uni <- rma(yi=logRR, sei=logRRse, slab=regions, measure="RR")

ci.exp(uni)

forest(uni, transf=exp, refline=1, pch=23, bg=4, col=2,
  main="Temperature and laid eggs")

```

Not so nice!

## 3. Lagged distributed case

Here we replicate the previous case, with lagged temperatures

Create empty matrices to store the COEF and VCOV

```{r}
coef <- matrix(NA,length(bio.list_sel), 4, dimnames=list(regions,paste0("b",seq(4))))
S <- vector("list",length(bio.list_sel))
names(S) <- regions

```

Define the crossbasis (CB) for temperature (fixed to define the same parametrization across regions)

```{r}
bound <- rowMeans(sapply(bio.list_sel, function(x) range(x$medianTweek, na.rm = T)))
varknots <- equalknots(bound, fun="bs", degree=2, df=4)
lagknots <- logknots(6, df=5, int=T) #modified lags from 21 to 6
argvar <- list(fun="bs", degree=2, knots=varknots, Bound=bound)
arglag <- list(fun="ns", knots=lagknots)


```

## 3.1 First stage

Run the loop

```{r}


for(i in seq(bio.list_sel)) {
  cb <- crossbasis(bio.list_sel[[i]]$medianTweek, lag=6, argvar=argvar, arglag=arglag)
  m <- glm(eggs ~ cb + ns(date, 2), family=quasipoisson(),
    data=bio.list_sel[[i]])
  cr <- crossreduce(cb,m, cen=3)
  coef[i,] <- coef(cr)
  S[[i]] <- vcov(cr)
}

```