---
title: "Two-stages time series analysis of *Aedes albopictus* oviposition"
author: "Daniele Da Re, Andrea Radici"
date: "2025-09-03"
output: 
  html_document:
    toc: true
    toc_depth: 4
    number_sections: false
    theme: readable
---


```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 

if (file.exists("C:/Users/2024ar003/Desktop/Alcuni file permanenti/Post_doc/Codice/local.R")){
opts_knit$set(root.dir = "C:/Users/2024ar003/Desktop/Alcuni file permanenti/Post_doc/Codice/04_Trento/GitShared_AlboLaggedAnalysis/")
} else {
opts_knit$set(root.dir = "/home/dared/GitHub/albopictus_lagged_analysis/")
}

```

## 0. Introduction

This report presents a time series analysis of *Aedes albopictus* egg abundance using the **VectAbundance dataset** ([Da Re et al., 2024](https://doi.org/10.1038/s41597-024-03482-y)). The study focuses on understanding the influence of climatic variables on egg dynamics, model validation, and climate change projections.

The specific objective of this study is to perform a two-stages analysis of the egg abundance temporal series to investigate the effect of lagged temperature and precipitation on egg-abundance analysis, first differentiating each regions and later pooling over the whole dataset.

## 1. Preprocessing

Load packages and observations. 

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(Epi) 
library(dlnm) 
library(tsModel)
library(splines) 
library(mgcv)
library(gnm) 
library(pbs) 
library(terra)
library(sf)
library(mapview)
library(metafor) 
library(scales)
library(lme4)
library(mixmeta)

bio.matrix_sel <- readRDS("VectAbundance015_ER5.RDS") %>% 
  select(ID, Country, Region, year, latitude, longitude, week, date, value, 
         tas, tp) %>% 
  rename(eggs=value, medianTweek= tas, cumPrecweek=tp, long= longitude, lat = latitude) %>% 
  mutate(month=month(date), 
         eggs=floor(eggs)) %>% 
  mutate(ID = as.factor(ID)) %>%
  as_tibble()
```
Remove the NAs

```{r}
bio.matrix_sel <- bio.matrix_sel %>% 
  filter(!is.na(eggs))
```

A map was created to illustrate the geographical distribution of the sampling locations.

```{r}
bio.matrix_sel %>% 
  select(long, lat) %>% 
  distinct() %>% 
  st_as_sf(coords=c("long", "lat"), crs=4326) %>% 
  mapview()
```
We apply log10 transformation to **brutally** account for the negative binomial distribution. we'll try not to use so far
```{r}
bio.matrix_sel$eggsL10<- log10(bio.matrix_sel$eggs+1)
```

## 2. Simple case (glm)

## 2.1 First stage

Following "Day5_Session1_Demo_TwoStage.R", we organise the data as a list of dataframes for the first stage. First, we define regions, supposedly homogeneous in terms of (a) climate (b) sampling (c) mosquito pressure... And we split our df as a list of df (one for each region)

```{r}

regions <- as.character(unique(bio.matrix_sel$Region))
bio.list_sel <- lapply(regions, function(r) bio.matrix_sel[bio.matrix_sel$Region==r,])
names(bio.list_sel) <- regions

```

We will first explore a simple, glm-based design, to obtain the relative risk as a function of temperature (but semi poisson). Natural spline over date and year, random effect over the ID.

```{r}

logRR <- logRRse <- vector("numeric", length(regions))
for(i in seq(bio.list_sel)) {
  
  cat(i, "\n")
  sub <- bio.list_sel[[i]]
  
  nYears = length(unique(sub$year))# to regulate the df of the ns as ns(date, nYears). 
  
  
  if(length(unique(sub$ID))>1){
    m <- glmer(eggs ~  medianTweek + ns(date, nYears) + (1|ID),
               data=sub, family=poisson())
    logRR[i] <- coef(m)$ID["medianTweek"][1,1]
    
    # m <- glm(eggs ~  medianTweek + ns(date, nYears),
    #          data=sub, family=poisson())
    # 
    # logRR[i] <- coef(m)["medianTweek"]
    
    # guarda AIC con e senza ID
    
    
  } else {
    m <- glm(eggs ~  medianTweek + ns(date, nYears),
             data=sub, family=poisson())
    
    logRR[i] <- coef(m)["medianTweek"]
  }
  
  logRRse[i] <- sqrt(vcov(m)["medianTweek","medianTweek"])
}


```

## 2.2 Second stage
Therefore we pool estimates (with univariate random effects meta-analysis) and we plot the RR per region and divided

```{r}

# uni <- rma(yi=logRR, sei=logRRse, slab=regions, measure="RR")
uni <- rma(yi=logRR, sei=logRRse, slab=regions, measure="GEN")
# sceglieire uno dei due

#

ci.exp(uni)

forest(uni, transf=exp, refline=1, pch=23, bg=4, col=2,
  main="Temperature and laid eggs")

```

This is nice eventually

## 3. Lagged distributed case

TEMPORARY Some regions give some problems later: remove them

Actually, at each time we remove a region, new regions create problems...


```{r}

regions <- as.character(unique(bio.matrix_sel$Region))
regions <- regions[!(regions %in% c("Vlore", "Autonomous Province of Trento", "Fier"))]
# regions <- regions[!(regions %in% c("Vlore", "Autonomous Province of Trento", "Sicily"))]
# regions <- regions[!(regions %in% c("Puglia", "Lushnje", "Fier", "Vlore", "Sicily", "Autonomous Province of Trento"))]
bio.list_sel <- lapply(regions, function(r) bio.matrix_sel[bio.matrix_sel$Region==r,])
names(bio.list_sel) <- regions

```

Here we replicate the previous case, with lagged temperatures. We create empty matrices to store the COEF and VCOV

```{r}
coef <- matrix(NA,length(bio.list_sel), 4, dimnames=list(regions,paste0("b",seq(4))))
S <- vector("list",length(bio.list_sel))
names(S) <- regions

```

Define the crossbasis (CB) for temperature (fixed to define the same parametrization across regions)

```{r}
# bound <- rowMeans(sapply(bio.list_sel, function(x) range(x$medianTweek, na.rm = T)))

bound <- c(0.14, 28.77) # force bounds to assume given values
varknots <- equalknots(bound, fun="bs", degree=2, df=4) #modified degree = 2, df = 4
lagknots <- logknots(6, df=5, int=T) #modified lags from 21 to 6
argvar <- list(fun="bs", degree=2, knots=varknots, Bound=bound)
arglag <- list(fun="ns", knots=lagknots)


```

## 3.1 First stage

Run the loop, which is quite slow

```{r}

for(i in seq(bio.list_sel)) {
  cat(i, "\n")
  cat(names(bio.list_sel[i]), "\n")
  sub <- bio.list_sel[[i]]
  
  cb <- crossbasis(sub$medianTweek, lag=6, argvar=argvar, arglag=arglag)
  # m <- glm(eggs ~ cb + ns(date, 2), family=quasipoisson(),
  #   data=sub)
  # 
  
  nYears = length(unique(sub$year))# to regulate the df of the ns as ns(date, nYears). 
  
  
  if(length(unique(sub$ID))>1){
    m <- glmer(eggs ~  cb + ns(date, nYears) + (1|ID),
               data=sub, family=poisson())
    
  } else {
    m <- glm(eggs ~ cb + ns(date, nYears),
             data=sub, family=poisson())
    
  }
  
  cr <- crossreduce(cb,m, cen=15)
  coef[i,] <- coef(cr)
  S[[i]] <- vcov(cr)
}

```

## 3.2 Second stage (Pooled)

### 3.2.1 Multivariate meta analysis

Therefore we should pool the estimates with mixmeta


```{r}

mix <- mixmeta(coef~1, S, method="ml")
print(summary(mix), digits=3)

```

### 3.2.2 Blups 

Best linear unbiased predictions (trade-off between location-specific estimates from the first stage and the pooled mean)

```{r}

mixblup <- mixmeta::blup(mix, vcov=TRUE)

```

### 3.2.3 pooled, first stage, blup estimates

Recompute spline transormation over weekly temperatures

```{r}

xvar <- seq(bound[1], bound[2], by=0.1)
bvar <- do.call("onebasis", c(list(x=xvar), argvar))
```


Compute predictions for the each estimate

```{r}
predpool <- crosspred(bvar, coef=coef(mix), vcov=vcov(mix), model.link="log",
  by=0.1, cen=15)
predreg <- lapply(seq(nrow(coef)),function(i) crosspred(bvar, coef=coef[i,],
  vcov=S[[i]], model.link="log", cen=15))
predblup <- lapply(seq(nrow(coef)),function(i) crosspred(bvar,
  coef=mixblup[[i]]$blup, vcov=mixblup[[i]]$vcov, model.link="log", cen=15))


```

And plot. So far, the predpool prediction is meaningless, but some of the predpool look more interesting.

```{r}

seq_plot = c(1:4, 6, 8, 9)

layout(t(1:2))
plot(predpool, type="l", ci="n", ylab="RR", ylim=c(.9,1.8), lwd=2,
  xlab="Average weekly temperature (°C)", main="Pooled and first-stage")
for(i in seq_plot) lines(predreg[[i]], col=alpha(4,0.3))

plot(predpool, type="l", ci="n", ylab="RR", ylim=c(.9,1.8), lwd=2,
  xlab="Average weekly temperature (°C)", main="Pooled and BLUPs")
for(i in seq_plot) lines(predblup[[i]], col=alpha(2,0.3))
layout(1)

```

